Marginal Models

### 2024-03-06 examples/_queue/_13_Examples_GrowthMixtureModels.qmd
config updates

---
title: "Multivariate Latent Growth Curve Models"
author: "Biostatistics Working Group"
---

```{r}
x <- 7;
x;
```

@@

--------

**STEP 1: Convert dataframe from long-to-wide format**
```{r long-to-wide}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Split the data into static (unchanging) and changing variables
# Static variables are taken only from the baseline measurement occasion
df_static <- df %>%
  filter(eventname == "baseline_year_1_arm_1") %>%
  select(src_subject_id, site_id_l, sex)

df_changing <- df %>%
  select(src_subject_id, eventname, anthroweightcalc, anthroheightcalc)

# Convert the changing variables from long format to wide format
df_wide_changing <- df_changing %>%
  gather(variable, value, -src_subject_id, -eventname) %>%
  unite("variable_event", variable, eventname) %>%
  spread(variable_event, value)

# Merge the static and changing data frames based on the 'src_subject_id' variable
df_wide <- df_static %>%
  left_join(df_wide_changing, by = "src_subject_id")

# Rename columns to shorter names for easier reference
df_wide <- df_wide %>% 
        rename("anthroweightcalc_0"="anthroweightcalc_baseline_year_1_arm_1",
               "anthroweightcalc_1"="anthroweightcalc_1_year_follow_up_y_arm_1",
               "anthroweightcalc_2"="anthroweightcalc_2_year_follow_up_y_arm_1",
               "anthroweightcalc_3"="anthroweightcalc_3_year_follow_up_y_arm_1",
               "anthroweightcalc_4"="anthroweightcalc_4_year_follow_up_y_arm_1",
               "anthroheightcalc_0"="anthroheightcalc_baseline_year_1_arm_1",
               "anthroheightcalc_1"="anthroheightcalc_1_year_follow_up_y_arm_1",
               "anthroheightcalc_2"="anthroheightcalc_2_year_follow_up_y_arm_1",
               "anthroheightcalc_3"="anthroheightcalc_3_year_follow_up_y_arm_1",
               "anthroheightcalc_4"="anthroheightcalc_4_year_follow_up_y_arm_1"
               )

# Reorder columns in the dataframe
df_wide <- df_wide %>% select(src_subject_id, site_id_l, sex, anthroweightcalc_0, anthroweightcalc_1, anthroweightcalc_2, anthroweightcalc_3, anthroweightcalc_4, anthroheightcalc_0, anthroheightcalc_1, anthroheightcalc_2, anthroheightcalc_3, anthroheightcalc_4)

The code fits a xxxx to examine the 'xxxxx' variable across time points ('eventname'). It incorporates xxxxxx. The results of the model are then printed to provide a detailed summary of the fitted model parameters.

**STEP 2: Compute xxxxx**

```{r}
# Specify the multivariate latent growth curve model
model <- '
  # Intercept and slope factors for variable 1
  i_var1 =~ 1*anthroheightcalc_0 + 1*anthroheightcalc_1 + 1*anthroheightcalc_2
  s_var1 =~ 0*anthroheightcalc_0 + 1*anthroheightcalc_1 + 2*anthroheightcalc_2

  # Intercept and slope factors for variable 2
  i_var1 =~ 1*anthroweightcalc_0 + 1*anthroweightcalc_1 + 1*anthroweightcalc_2
  s_var1 =~ 0*anthroweightcalc_0 + 1*anthroweightcalc_1 + 2*anthroweightcalc_2
'
```

>>>>

---
title: "Growth Mixture Models"
author: "Biostatistics Working Group"
---

@@

**STEP 1: XXXXXXXXX **

----

```{r}
# Combine the data from all classes
data <- rbind(class1, class2, class3)
```

### Model Specification and Estimation
Now that we have our simulated data, we can fit a Growth Mixture Model. In this example, we will assume that there are three latent classes. However, in practice, the number of classes is usually unknown and must be determined through model comparisons or other methods.
```{r, eval=FALSE}
# Fit the Growth Mixture Model
gmm <- hlme(value ~ time,
            random = ~ time,
            subject = "id",
            mixture = ~ time,
            data = data,
            ng = 3,
            B = list(value = c(1, 0, 0, 0)
```

### 2024-03-06 examples/_queue/_15_Examples_RandomInterceptCrosslaggedPanelModels.qmd
config updates

---
title: "Multivariate Latent Growth Curve Models"
author: "Biostatistics Working Group"
---

```{r}
x <- 7;
x;
```

@@

```{r}
# Specify the multivariate latent growth curve model
model <- '
  # Intercept and slope factors for variable 1
  i_var1 =~ 1*anthroheightcalc_0 + 1*anthroheightcalc_1 + 1*anthroheightcalc_2
  s_var1 =~ 0*anthroheightcalc_0 + 1*anthroheightcalc_1 + 2*anthroheightcalc_2

  # Intercept and slope factors for variable 2
  i_var1 =~ 1*anthroweightcalc_0 + 1*anthroweightcalc_1 + 1*anthroweightcalc_2
  s_var1 =~ 0*anthroweightcalc_0 + 1*anthroweightcalc_1 + 2*anthroweightcalc_2
'
```

```{r}
# Estimate the model
fit <- lavaan::sem(model, data = df_wide, missing = "FIML")

>>>>

---
title: "Random-Intercept Crosslagged Panel Models"
author: "Biostatistics Working Group"
---

@@

```{r, eval=FALSE}
# Define the model
model <- '
  # Random intercepts
  i_X =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2 + 1 * anthroheightcalc_3 + 1 * anthroheightcalc_4
  i_Y =~ 1 * anthroweightcalc_0 + 1 * anthroweightcalc_1 + 1 * anthroweightcalc_2 + 1 * anthroweightcalc_3 + 1 * anthroweightcalc_4

  # Cross-lagged effects
  anthroweightcalc_1 ~ beta_YX * anthroheightcalc_0
  anthroweightcalc_2 ~ beta_YX * anthroheightcalc_1
  anthroweightcalc_3 ~ beta_YX * anthroheightcalc_2
  anthroweightcalc_4 ~ beta_YX * anthroheightcalc_3

  anthroheightcalc_1 ~ beta_XY * anthroweightcalc_0
  anthroheightcalc_2 ~ beta_XY * anthroweightcalc_1
  anthroheightcalc_3 ~ beta_XY * anthroweightcalc_2
  anthroheightcalc_4 ~ beta_XY * anthroweightcalc_3

  # Auto-regressive paths (optional but often included in CLPM)
  anthroheightcalc_1 ~ alpha_X * anthroheightcalc_0
  anthroheightcalc_2 ~ alpha_X * anthroheightcalc_1
  anthroheightcalc_3 ~ alpha_X * anthroheightcalc_2
  anthroheightcalc_4 ~ alpha_X * anthroheightcalc_3

  anthroweightcalc_1 ~ alpha_Y * anthroweightcalc_0
  anthroweightcalc_2 ~ alpha_Y * anthroweightcalc_1
  anthroweightcalc_3 ~ alpha_Y * anthroweightcalc_2
  anthroweightcalc_4 ~ alpha_Y * anthroweightcalc_3

  # Covariances (optional but can be included for model fit)
  anthroheightcalc_0 ~~ anthroweightcalc_0
  i_X ~~ i_Y
  '
```

```{r}
# Estimate the RI-CLPM 
fit <- sem(model, data = df_wide, missing = "FIML")

### 2024-03-06 examples/_queue/_16_Examples_LatentCurveModelsStructuredResiduals.qmd
config updates

---
title: "Latent Growth Curve Models"
author: "Biostatistics Working Group"
---

@@

# Define the Latent Growth Curve Model
model <- '
  # Latent variables
    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2

  # Estimate means of latent variables
    i ~ 1
    s ~ 1

  # Variances
    i ~~ i
    s ~~ s

  # Covariance
    i ~~ s
'
```

```{r}
# Fit the model
fit <- sem(model, data = df_wide, missing = "FIML")

>>>>

---
title: "Latent Curve Models with Structured Residuals"
author: "Biostatistics Working Group"
---

@@

model <- '
  # Latent variables
  intercept =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2 + 1 * anthroheightcalc_3 + 1 * anthroheightcalc_4
  slope =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2 + 3 * anthroheightcalc_3 + 4 * anthroheightcalc_4

  # Time-varying covariate effect on outcome
  anthroheightcalc_0 ~ c1 * anthroweightcalc_0
  anthroheightcalc_1 ~ c2 * anthroweightcalc_1
  anthroheightcalc_2 ~ c3 * anthroweightcalc_2
  anthroheightcalc_3 ~ c4 * anthroweightcalc_3
  anthroheightcalc_4 ~ c5 * anthroweightcalc_4

  # Residual variances and covariances
  anthroheightcalc_0 ~~ r1 * anthroheightcalc_0
  anthroheightcalc_1 ~~ r2 * anthroheightcalc_1
  anthroheightcalc_2 ~~ r3 * anthroheightcalc_2
  anthroheightcalc_3 ~~ r4 * anthroheightcalc_3
  anthroheightcalc_4 ~~ r5 * anthroheightcalc_4

  # Latent variable covariances
  intercept ~~ i_s_cov * slope
'

```{r}
# Estimate the LCM-SR Model
fit <- sem(model, data = df_wide, missing = "FIML")

### 2024-03-06 examples/_queue/_5_Examples_MarginalModels.qmd
config updates

---
title: "Linear Mixed Models: Random Intercept and Slope"
author: "Biostatistics Working Group"

---

# LMM: Intercept & Slope {.unnumbered}

@@

The Linear Mixed Model with a random intercept and slope (LMM:ris) extends the traditional fixed-effect linear regression by incorporating both a subject-specific random intercept and a random slope. This allows each participant to have their own unique intercept and slope values, reflecting individual starting points and rates of change, in addition to the overall mean-level (fixed-effect) trajectory.

In this example, we will utilize the LMM:ris to analyze height trajectories obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study. Our primary objective is to understand the stability and evolution in height measurements, while factoring in the clustered nature of observations within individuals over time. The LMM:ris facilitates this by concurrently modeling an overarching sample mean trajectory (fixed effect) and individual variations (random effects) in both starting points (intercepts) and growth rates (slopes) around this mean trajectory.

@@

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT","lme4","report","broom","gridExtra","easystats","gtsummary")

@@

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables
library(lme4)         # Linear mixed-effects models
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page
library(easystats)    #
library(gtsummary)    #
```
:::

@@

The code fits a linear mixed model to examine the 'Height' variable across time points ('eventname'). It incorporates both random intercepts and slopes for the time points ('eventname') within each participant ('src_subject_id') to capture individual-level variability in both starting values and rates of change over time. The results of the model are then printed to provide a detailed summary of the fitted model parameters.

**STEP 1: Compute LMM with Random Intercepts and Slopes**
```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

## Linear Mixed Model with a random intercept and random slope (LMM-ris)
random_intercepts_slopes <- lmer(anthroheightcalc ~ 1 + eventname + (1|src_subject_id), data = df, REML=T)

print(random_intercepts_slopes)

```

```{r model-output}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

## Output and reports extending from the LMM-ris analyses
summary(random_intercepts_slopes)
confint(random_intercepts_slopes, level = 0.95, method = "Wald")
report_performance(random_intercepts_slopes)

```

The code fits a Linear Mixed Model (LMM:ris) to predict children's height based on different time points (denoted as eventname). Individual variability is captured through random intercepts associated with each participant (src_subject_id). The results suggest a significant change in height across the time points, with the model accounting for 89% (r round(conditional_R2, 2)) of the total variation and 37% (r round(marginal_R2, 2)) attributed to the fixed effects alone.

In the model, the linear term for eventname has an estimated effect of r round(model_summary$coefficients["eventname.L", "Estimate"], 2), indicating a significant increase in height across time points. Additionally, there's variability in initial height across participants, reflected by a random intercept standard deviation of r round(random_effects["src_subject_id", "Std.Dev."], 2).

### Model Plots

The following set of plots are used to facilitate model diagnostics. The first is a histogram showcasing the distribution of random intercepts for individual subjects, indicating variations in height not explained by the fixed effects. The second depicts residuals versus fitted values, helping assess the model's fit and potential heteroscedasticity. The third contrasts observed and predicted height values across different time points, offering a side-by-side evaluation of the model's predictions against actual observations.

::: blue
```{r diagnosticplots}

# Assuming your model is named `random_intercepts_slope`
# 1. Extract the random effects
random_effects <- ranef(random_intercepts_slopes)[[1]]

# 2. Convert to dataframe
random_effects_df <- data.frame(Intercept = random_effects$`(Intercept)`)

# Plot 1: Histogram
# Plot 1: Histogram
hist_plot <- ggplot(random_effects_df, aes(x = Intercept)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
  theme_minimal()

print(hist_plot)

```

```{r trajectoryplot}

# Extract the data frame used in the model
model_data <- random_intercepts_slopes@frame

# Extract unique subject IDs from the model's data
original_subject_ids <- unique(model_data$src_subject_id)

# Subset the original data to include only those subjects
df_subset <- df %>% filter(src_subject_id %in% original_subject_ids)

eventname_map <- c(
  "baseline_year_1_arm_1" = "Baseline",
  "1_year_follow_up_y_arm_1" = "Year_1",
  "2_year_follow_up_y_arm_1" = "Year_2",
  "3_year_follow_up_y_arm_1" = "Year_3",
  "4_year_follow_up_y_arm_1" = "Year_4"
)

# Plot
ggplot(df, aes(x = eventname, y = anthroheightcalc, group = src_subject_id)) +

# Individual estimated height trajectories in faded lines
geom_line(aes(group = src_subject_id), alpha = 0.3, color = "grey50") +

# Overall group-mean trajectory in blue with increased thickness
stat_summary(aes(group = 1), fun = mean, geom = "line", color = "blue", linewidth = 1) +
labs(title = "Individual and Group-Mean Height Trajectories",
     x = "Event Name",
     y = "Height") +
  theme_minimal()

```

The provided code visualizes individual and group-mean height trajectories over different event names. Individual height trajectories for each subject are depicted as faded gray lines, allowing for a clear view of the variability among subjects. In contrast, the overall group-mean trajectory, which represents the average trend across all individuals for each event name, is highlighted in blue. The average height at r mean(df_descriptable$Height) has increased by r mean(df_descriptable$Height_followup - df_descriptable$Height) units from baseline.

## Wrapping Up
<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

The linear mixed model analysis was conducted to predict children's height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4). The eventname predictor was statistically significant with a p-value of r format.pval(fixed_effects$Pr(>|t|)[2], digits = 3). The model's overall ability to explain the variance in height was xxxxx, with a conditional R^2 of r report_performance(random_intercepts)$Conditional R2[1], indicating that it accounted for this proportion of the variability in height when considering both fixed and random effects. The marginal R^2 was r report_performance(random_intercepts)$Marginal R2[1], meaning that the fixed effects alone explained this proportion of the variability.

>>>>

---
title: "Marginal Models"
author: "Biostatistics Working Group"
---

# Marginal Models {.unnumbered}

@@

Marginal models xxxxxxx. This xxxxxxx.

In this example, we will utilize marginal models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study. Our primary objective is to understand xxxxxx, while factoring in the clustered nature of observations within individuals over time. Marginal models facilitate this by xxxxx.

@@

# Create a list of required packages
packages_required <- c("tidyverse","lme4","geepack", "report",
                       "broom","gridExtra","easystats","gtsummary",
                       "marginaleffects")

@@

library(tidyverse)    # Collection of R packages for data science
library(lme4)         # Linear mixed-effects models
library(geepack)      # Generalized Estimating Equations (GEE) analysis
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page
library(easystats)    #
library(gtsummary)    #
library(marginaleffects) #
```
:::

@@

The code fits a xxxx to examine the 'xxxxx' variable across time points ('eventname'). It incorporates xxxxxx. The results of the model are then printed to provide a detailed summary of the fitted model parameters.

**STEP 1: Compute xxxxx**
```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Fitting the longitudinal marginal model
model <- geeglm(anthroweightcalc ~ eventname, data = df, id = src_subject_id, family = gaussian(), corstr = "exchangeable")
#model <- geeglm(Job_Satisfaction ~ TimePoint, data = data, id = Individual, family = #binomial(link = "logit"), corstr = "exchangeable")
```

```{r}
summary(model)
```
